---
title: "The Isomorphism Between Ant Colonies and Ensemble Learning"
author: "Ernest Fokoué, Gregory Babbitt, Yuval Levental"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Isomorphism Between Ant Colonies and Ensemble Learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5, fig.align = "center"
)
library(IsomorphismSim)
set.seed(42)
```

## Overview

This vignette demonstrates the core claim of the paper: **ant colony
decision-making and random forest learning are isomorphic** under a common
formalism of stochastic ensemble intelligence. Both systems reduce variance
through decorrelation of identical units, and the governing variance
decomposition is:

$$
\operatorname{Var}[\text{ensemble}] = \rho\,\sigma^2 + \frac{1-\rho}{M}\,\sigma^2
$$

where $\rho$ is the pairwise correlation between units, $\sigma^2$ is the
individual-unit variance, and $M$ is the ensemble size.

## 1. Ant Colony Simulation

The `simulate_ant_colony()` function runs an agent-based simulation of
*Temnothorax*-style nest-site selection. The key parameter is `p_explore`,
which controls how often ants explore randomly (decorrelation) vs follow
pheromone trails (exploitation).

```{r colony}
sim <- simulate_ant_colony(
  n_ants    = 50,
  p_explore = 0.3,
  n_steps   = 100,
  noise_sd  = 2
)
cat("Colony chose site", sim$decision,
    "(true best: 1)\n")
cat("Colony quality estimates:",
    sprintf("%.1f", sim$final_preferences), "\n")
```

## 2. The Correlation-Decorrelation Trade-off

Running many colonies at different `p_explore` values shows that correlation
between ant assessments decreases as exploration increases — exactly as
tree correlation decreases with lower `m_try/p` in random forests.

```{r sweep, fig.cap="Correlation decay mirrors the theoretical prediction.", cache=TRUE}
thetas <- c(0.1, 0.3, 0.5, 0.7, 0.9)
n_reps <- 20
results <- data.frame()

for (theta in thetas) {
  pmat <- matrix(NA, 50, n_reps)
  for (r in seq_len(n_reps)) {
    s <- simulate_ant_colony(n_ants = 50, p_explore = theta)
    pmat[, r] <- s$ant_preferences[, 1]
  }
  cm <- cor(pmat, use = "pairwise.complete.obs")
  results <- rbind(results, data.frame(
    theta = theta,
    rho   = mean(cm[lower.tri(cm)], na.rm = TRUE)
  ))
}

library(ggplot2)
rho_max <- max(results$rho)
theory <- data.frame(
  theta = seq(0, 1, 0.01),
  rho   = rho_max * (1 - seq(0, 1, 0.01))
)

ggplot() +
  geom_line(data = theory, aes(theta, rho),
            linewidth = 1, linetype = "solid") +
  geom_point(data = results, aes(theta, rho),
             color = "#BF360C", size = 3) +
  geom_line(data = results, aes(theta, rho),
            color = "#BF360C", linetype = "dashed") +
  labs(title = "Ant Colony: Correlation Decay",
       x = expression(theta ~ "(p_explore)"),
       y = expression(rho)) +
  theme_minimal(base_size = 13)
```

## 3. Variance Decomposition Demo

The theoretical variance formula can be visualised directly:

```{r variance_curve, fig.cap="Ensemble variance converges to the irreducible floor."}
rho <- 0.4; sigma2 <- 1.5
M <- 1:300
v <- rho * sigma2 + (1 - rho) * sigma2 / M

plot(M, v, type = "l", lwd = 2, col = "#1565C0",
     xlab = "Ensemble size M", ylab = "Var[ensemble]",
     main = expression("Var = " * rho * sigma^2 +
                         (1 - rho) * sigma^2 / M))
abline(h = rho * sigma2, lty = 2, col = "red")
legend("topright",
       legend = c("Ensemble variance",
                  bquote(rho * sigma^2 == .(rho * sigma2))),
       col = c("#1565C0", "red"), lty = c(1, 2), lwd = 2)
```

## 4. Running the Full Experiment Suite

The package provides functions for all five experiments from the paper.
For a quick demonstration (reduced replicates):

```r
# Experiment 1: RF variance decomposition
rf_res <- variance_decomposition_experiment(
  n_replicates = 10, n_trees = 100, n_test = 200
)
plot_variance_decomposition(rf_res)

# Experiment 2: Colony variance decomposition
col_res <- colony_variance_experiment(
  n_ants_values = c(10, 50), n_replicates = 10
)
plot_colony_accuracy(col_res)

# Experiment 3: Direct isomorphism test
iso_res <- isomorphism_test(n_replicates = 10)
plot_correlation_decay(iso_res)
```

## 5. Interactive Exploration

Launch the Shiny app for a richer, interactive experience:

```r
launch_app()
```

The app provides three tabs:

- **Colony Simulation**: Run a single colony with adjustable parameters and
  watch pheromone/recruitment dynamics unfold.
- **Isomorphism Sweep**: Sweep `p_explore` and see the correlation decay curve
  emerge in real time.
- **Variance Decomposition**: An interactive demo of the variance formula.

## Reference

Fokoué, E., Babbitt, G., & Levental, Y. (2026). Decorrelation, Diversity,
and Emergent Intelligence: The Isomorphism Between Social Insect Colonies
and Ensemble Machine Learning. *Rochester Institute of Technology*.
